---
title: "geosR Tutorial: Resource Estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{geosR_tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The `geosR` package provides a generalized geostatistical workflow to calculate spatial resources (volume, metal/content tonnage, and average grades). It simplifies the typical `gstat` + `sf` pipeline utilized by spatial geologists.

There are 5 core functions:
1. `fit_var()`: Fit a variogram model to spatial drillhole data.
2. `est_krige()`: Run an ordinary kriging block model prediction.
3. `calc_res()`: Intersect block model results with an area of interest polygon to compute volume, grades, and tonnage.
4. `ev_rest()`: Perform resource reconciliation by comparing calculated totals against actual production realization.
5. `plot_res()`: Visually plot the spatial resource distributions across your blocks.

```{r setup}
library(geosR)
library(sf)
library(terra)
```

## Basic Workflow Example

Below is a theoretical workflow implementing these functions.

### 1. Load Data
Assume you have loaded a spatial `sf` object `drill_data` and an `area` boundary polygon.

```R
# drill_data <- st_as_sf(df, coords=c("X","Y"), crs=32748)
```

### 2. Variogram Modeling
Fit a spherical model to your spatial grade data:
```R
var_model <- fit_var(
  data = drill_data, 
  formula = grade ~ 1, 
  cutoff = 250, 
  width = 50, 
  model_type = "Sph",
  sill = 0.02, 
  range = 150
)

# Plot the model
plot(var_model$variogram, var_model$model, main="Spherical Model")
```

### 3. Ordinary Kriging
Generate a calculation grid, and predict the variable utilizing the fitted model.
```R
calc_grid <- st_make_grid(st_bbox(area), cellsize = 20)

kriged_grade <- est_krige(
  data = drill_data, 
  formula = grade ~ 1,
  grid = calc_grid, 
  vgm_model = var_model$model
)
```

### 4. Calculate Resources
Compute your overall volume and tonnage inside an area parameter.
```R
# Convert kriged points to rasters for both grade and thickness
raster_grade <- as(st_rasterize(kriged_grade["var1.pred"]), "SpatRaster")
raster_thick <- as(st_rasterize(kriged_thick["var1.pred"]), "SpatRaster")

# Density determines your resulting tonnage unit (assuming grade is raw % or kg/m3)
my_resources <- calc_res(raster_grade, raster_thick, area, density = 1.0)

# Check calculated figures per Area Polygon
print(my_resources$table)
```

### 5. Evaluate Realization
Determine a recovery factor by contrasting with actual plant performance.
```R
final_table <- ev_rest(my_resources$table, actual_production = 83000)
print(final_table)
```

### 6. Plot the Estimate
Generate standard plots to add to reports.
```R
plot_res(
  tonnage_raster = my_resources$raster, 
  area = area, 
  title = "Resource Estimate Block A"
)
```
